$K$-corrections and filter transformations \\ from the restframe ultraviolet to the near infrared

### 2006-04-19 docs/paper/kcorrect.tex
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@550 75dff457-8751-0410-ab1a-cd701654ce6b

\documentclass[10pt,preprint]{aastex}

@@

@@

In this section, we describe how to take the astrophysical models and
the astronomical observations, and to pose the problem in the form
needed for the purposes of NMF. Section \ref{models} describes the
basis set of models, Section \ref{In Appendix \ref{nmf} we describe
how to solve to NMF problem.

@@

[sam sam sam sam]

>>>>

\documentclass[10pt,preprint]{aastex}
\usepackage{amsmath}
\usepackage{epsfig}

@@

\def\beq#1\eeq{\begin{equation}#1\end{equation}}
\def\beqa#1\eeqa{\begin{align}#1\end{align}}
\def\beqq#1\eeqq{\begin{equation*}#1\end{equation*}}
\def\beqqa#1\eeqqa{\begin{align*}#1\end{align*}}
\def\bseq#1\eseq{\begin{subequations}#1\end{subequations}}

\newcommand{\T}{^{\scriptscriptstyle \top}}
\newcommand{\iT}{^{\scriptscriptstyle -\top}}
\newcommand{\inv}{^{-1}}

\newcommand{\ie}{i.e.}

\newcommand{\Log}{\mathrm{Log}}
\newcommand{\trace}{\mathrm{Tr}}
\newcommand{\detM}{\det(M)}

\newcommand{\XX}{X}
\newcommand{\XXt}{\tilde{X}}
\newcommand{\XXh}{\hat{X}}
\newcommand{\Xij}{X_{ij}}
\newcommand{\Xijh}{\hat{X}_{ij}}
\renewcommand{\AA}{A}
\newcommand{\AAt}{\tilde{A}}
\newcommand{\WW}{W}
\newcommand{\WWt}{\tilde{W}}
\newcommand{\Wik}{W_{ik}}
\newcommand{\HH}{H}
\newcommand{\HHt}{\tilde{H}}
\newcommand{\Hkj}{H_{kj}}
\newcommand{\BB}{B}
\newcommand{\BBt}{\tilde{B}}
\newcommand{\MM}{M}
\newcommand{\MMt}{\tilde{M}}
\newcommand{\Aik}{A_{ik}}
\newcommand{\Bkr}{B_{kr}}
\newcommand{\Mrj}{M_{rj}}
\newcommand{\sij}{\sigma_{ij}}

@@

In this section, we describe how to take the astrophysical models and
the astronomical observations, and to pose the problem in the form
needed for the purposes of NMF. Section \ref{models} describes the
basis set of models, Section {\bf redo} In Appendix \ref{nmf} we describe
how to solve to NMF problem.

@@

\subsection{Standard NMF}
The standard non-negative matrix factorization (NMF) problem, as
originally posed by Lee and Seung\cite{??}, is to approximate a data
matrix $\XX$ (of size $n\times d$) as the outer product of two
rank-$k$ non-negative matrices $\WW$ and $\HH$. For example, in this
paper, $\XX$ represents the fluxes at $n$ different wavelengths for
$d$ different galaxies, $\HH$ represents a set of templates to build
each galaxy from, and $\WW$ represents, for each galaxy, the weights
to give each template. This factorization is similar in spirit to
singular value decomposition (SVD) and principal components analysis
(PCA) but crucially, in NMF, $\XX$, $\WW$ and $\HH$ are all
constrained to have non-negative entries. This changes the problem
fundamentally, since no ``cancellation'' of positive and negative
basis functions or coefficients is possible -- all approximation
interactions are strictly additive. The goal, as with SVD or PCA, is
to minmize the squared approximation error (Frobenius norm):
\beqa
\chi^2 &= \|\XX-\WW\HH\|^2\\
%&= \|\XX-\XXh\|^2\\
&= \sum_{ij} \left(\Xij - \sum_k \Wik\Hkj \right)^2
\eeqa
In their seminal papers\cite{??}, Lee and Seung show that this approximation
error in non-increasing under the following very simple multiplicative
update rules: 
\beqa
\Wik &\leftarrow \Wik \frac{[\XX\HH\T]_{ik}}{[\WW\HH\HH\T]_{ik}}\\
\Hkj &\leftarrow \Hkj \frac{[\WW\T\XX]_{kj}}{[\WW\T\WW\HH]_{kj}}
\eeqa
$[M]_{ab}$ denotes the $(ab)$ element of the matrix valued expression
$M$. $\HH\T$ is the transpose of matrix $\HH$.

These rules are remarkable because, although they make finite (not
infinitesimal) adjustments to the elements of the approximation
matrices, they have no step-size parameters and are always guaranteed
to reduce the error (or leave it invariant once they have converged).
$\chi^2$ is in general a ``non-convex'' function of $\WW$ and $\HH$,
meaning we cannot guarantee there is only one local
minimum. Therefore, the procedure does not necessarily find the global
optimum. But in practice, even with random initialization, these rules
tend to converge to good solutions when the data matrix $\XX$ is
highly structured. {\bf explain!}

\subsection{NMF in the space of coefficients for a known basis}
The factorization problem we face in this paper is a slight variation on the
basic NMF setup described above. We wish to approximate $\XX$ by the
product $\AA\BB\MM$, 
where $\MM$ is a given (fixed) basis matrix of size $r \times d$
and the optimization is over the matrices $\AA$ (of size
$n\times k$) and $\BB$ (of size $k\times r$). All matrices
$\XX,\AA,\BB\,\MM$ have non-negative entries. This can be thought of
as performing non-negative matrix factorization on the coefficients of
an approximation, given a fixed (non-negative) basis, described by the
columns of $\MM$. Once again, the objective we wish to minimize is the
squared approximation error: 
\beqa
\chi^2 &= \|\XX-\AA\BB\MM\|^2\\
&= \sum_{ij} \left(\Xij - \sum_{kr} \Aik\Bkr\Mrj \right)^2
\eeqa
In the case of this paper, this generalization allows us to express
the templates in terms of star-formation histories, but to compare the
predicted fluxes for the star-formation histories through the $\MM$
matrix to the observed fluxes $\XX$.

Of course, in the very special case where $\MM$ is an invertible
matrix, this problem can be transformed into the original NMF problem above by
right multiplying the data matrix by $\MM\inv$. However, in most
situations, including ours, $\MM$ has many fewer rows than columns, and
as such is far from invertible. Fortunately, however, it is possible
to derive multiplicative updates for this extended problem 
which minimze the error directly, even when $\MM$ is not invertible.

First, by thinking of the tensor product $\BB\MM$ as a single
non-negative matrix $\HH$, we can trivially derive a multiplicative
update equation for the elements of $\AA$ by using the $\WW$ update
provided above: 
\beq
\Aik \leftarrow \Aik 
\frac{[\XX\MM\T\BB\T]_{ik}}{[\AA\BB\MM\MM\T\BB\T]_{ik}}
\eeq
By Lee and Seung's original proof, this update is guaranteed (for any
non-negative matrices $\XX$,$\MM$ and $\BB$) not to increase the
approximation error.

Our main algorithmic contribution is to derive a similar update equation
for the elements of $\BB$:
\beq
\Bkr \leftarrow \Bkr
\frac{[\AA\T\XX\MM\T]_{kr}}{[\AA\T\AA\BB\MM\MM\T]_{kr}}
\eeq
Following the technique outlined in \cite{??}, it can still be shown
that the error is nonincreasing under the application of this update.

\subsection{Nonuniform uncertainties}
When different entries $\Xij$ of the data matrix have different
uncertainties, the natural objective function is the weighted approximation
error (which also corresponds to the negative log likelihood under a
Gaussian noise assumption):
\beq
\chi^2 = \sum_{ij} \left( \frac{\Xij - 
\sum_k \Wik\Hkj}{\sij}\right)^2
\eeq

This case can easily be handled since during the update for $\WW$ the
elements of $\HH$ are fixed and vice versa and the updates are
guaranteed not to increase the cost for any nonnegative matrices.
%In particular, define $\XXt_{ij}=\Xij/\sij$,
%$\WWt(j)_{ik}=\Wik/\sij$ and 
%$\HHt(i)_{kj}=\Hkj/\sij$. 
In particular, when updating $\WW$, we can rewrite the cost function as
\beq
\chi^2 = \sum_{ij} \left( \frac{\Xij}{\sij} - 
\sum_k \Wik\frac{\Hkj}{\sij}\right)^2
%= \sum_{ij} \left(\XXt_{ij} - 
%\sum_k \Wik\HHt(i)_{kj}\right)^2
\eeq
yielding updates of the form:
\beq
\Wik \leftarrow \Wik 
\left( \sum_j \frac{\Xij\Hkj}{\sij^2} \right) / 
\left( \sum_{mj} \frac{W_{im}H_{mj}\Hkj}{\sij^2}\right)
%\Wik \leftarrow \Wik \frac{[\XXt\HHt(i)\T]_{ik}}{[\WW\HHt(i)\HHt(i)\T]_{ik}}
\eeq

and similarly, when updating $\HH$, we can rewrite the cost function as
\beq
\chi^2 = \sum_{ij} \left( \frac{\Xij}{\sij} - 
\sum_k \frac{\Wik}{\sij}\Hkj\right)^2
%= \sum_{ij} \left(\XXt_{ij} - 
%\sum_k \WWt(j)_{ik}\Hkj\right)^2
\eeq
yielding updates of the form:
\beq
\Hkj \leftarrow \Hkj
\left( \sum_i \frac{\Wik\Xij}{\sij^2}\right) /
\left( \sum_{mi} \frac{\Wik{}W_{im}H_{mj}}{\sij^2}\right)
%\Hkj \leftarrow \Hkj \frac{[\WWt(j)\T\XX]_{kj}}{[\WWt(j)\T\WWt(j)\HH]_{kj}}
\eeq

This argument can be equally applied to our extended model, yeilding
the final update equations which are actually implemented in the
kcorrect code:
\centerline{\fbox{\parbox{4.3in}{
\beqa 
\Aik &\leftarrow \Aik 
\left( \sum_{rj} \frac{\Xij\Mrj\Bkr}{\sij^2}\right) / 
\left( \sum_{mnrj} \frac{A_{im}B_{mr}M_{rj}M_{jn}B_{kn}}{\sij^2}\right) \\
\Bkr &\leftarrow \Bkr
\left( \sum_{ij} \frac{\Aik\Xij\Mrj}{\sij^2}\right) / 
\left( \sum_{mnij} \frac{A_{ik}A_{im}B_{mn}M_{nj}M_{rj}}{\sij^2}\right)
\eeqa
}}}

%\centerline{\fbox{\parbox{2.6in}{
%\beqa
%\Aik &\leftarrow \Aik 
%\frac{[\XXt\MMt(i)\T\BB\T]_{ik}}{[\AA\BB\MMt(i)\MMt(i)\T\BB\T]_{ik}}\\
%\Bkr &\leftarrow \Bkr
%\frac{[\AAt(?)\T\AAt(?)\BB\MM\MM\T]_{kr}}{[\AAt(?)\T\XXt\MM\T]_{kr}}
%\eeqa
%}~~}}\\[.2in]
%pwhere $\XXt_{ij}=\Xij/\sij$, $\MMt(i)_{rj} = \Mrj/\sij$ and
%$\AAt(?)_{ik}=\Aik/??$. 
%}}}
